---
title: 'Re-analysis of Carballo (2022) co-phylogenetic plot'
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
author: Liam J. Revell
---

As I mentioned in the Slack, I'm presently working on a new (extend) application note for the updated *phytools* R package. In this article I have included a number of different, fully-functional R workflows, similar to those that we use in our class exercises and that fill my book with Luke. *phytools* contains a very popular co-phylogenetic ("tanglegram") plotting method called `cophylo`. Unfortunately, I was having a lot of difficulty encountering published data to illustrate it in my paper!

I found one example that I liked in a paper by Caraballo ([2022](https://doi.org/10.1128/spectrum.01411-22)), giving the phylogenetic association of bats and their betacoronaviruses. Unfortunately, the author didn't publish the trees or original data in a usable format. The only "available" data was a set of tables (published in PDF format, no less) with *GenBank* accession numbers for all of the sequences used in the study!

I decided to start there anyway, and see what I could do to pull down each of these sequences based on their accession numbers, align the sequences, and do phylogenetic inference in R.

As an important sidenote, to replicate this analysis you will need to download the stand-alone program *MUSCLE* ([here](https://www.drive5.com/muscle/)), and then (in Windows) add the directory of this program to your [PATH environment variable](https://geekrewind.com/how-to-add-and-edit-path-environment-variables-in-windows-11/). Other than that, I just pulled the PDFs from Caraballo's supplement, used Adobe to convert it to a Word doc, imported to Excel, and saved as a CSV file to read into R!

Load packages.

```{r}
library(phytools)
library(phangorn)
```

Read my bat accession data.

```{r}
Bats<-read.csv(file="Bats.csv")
Bats
```

This contains accession numbers for both alpha- and betacoronaviruses. I decided to just do the betacoronaviruses.

```{r}
betaAccessions<-setNames(
	Bats$Accession[which(Bats$Host.type=="betaCoV")],
	Bats$Species[which(Bats$Host.type=="betaCoV")])
betaAccessions
```

Next, I'll use the *ape* function `read.GenBank` to pull the corresponding sequences from *GenBank*.

```{r}
betaBats<-read.GenBank(betaAccessions,seq.names=names(betaAccesions))
betaBats
```

These sequences are all *almost* the same length, so they may already be (mostly) aligned. We can use MUSCLE to align them anyway.

```{r}
betaBats.aligned<-muscle5(betaBats)
betaBats.aligned
```

We're already ready to estimate our tree. We could do model selection, but I decided just to use `"GTR+G(4)+I"`.

```{r}
betaBat_pml<-pml_bb(betaBats.aligned,model="GTR+G(4)+I")
```

Done! Let me pull out my tree.

```{r}
bat_mltree_unrooted<-betaBat_pml$tree
bat_mltree_unrooted
```

We can proceed to root it. Generally speaking, this wouldn't be advised, but I happen to know that the method of *midpoint rooting* will give me the tree of Caraballo (2022).

```{r}
plotTree(bat.tree<-untangle(midpoint(bat_mltree_unrooted),"read.tree"))
```

This tree has accession numbers instead of species labels -- but we know how they're related, so let's update them.

```{r}
spp<-setNames(names(betaAccessions),betaAccessions)
bat.tree$tip.label<-spp[bat.tree$tip.label]
plotTree(bat.tree,ftype="i")
```

Great. So far, so good.

Next, let's do much the same with our betacoronavirus sequences. In this case, we don't have named lineages -- all we've got are the *GenBank* labels, so we can just keep these as our OTU names.

```{r}
betaCovs<-read.csv(file="betaCovs.csv")
betaCovs_seqs<-read.GenBank(betaCovs$Accession.Number)
betaCovs_seqs.aligned<-muscle5(betaCovs_seqs)
betaCovs_tree<-pml_bb(betaCovs_seqs.aligned,
  model="GTR+G(4)+I")
plotTree(virus.tree<-untangle(midpoint(betaCovs_tree$tree),
  "read.tree"))
```

We can next build a table of associations between the taxa in our bat tree and the taxa in our virus tree. Once again, this is from the supplemental material of Caraballo (2022).

```{r}
assoc<-data.frame(Bats=betaCovs$Host.species,
  betaCoVs=betaCovs$Accession.Number)
assoc
```

One thing I noticed right away is that the tip labeled `"Glossophaga soricina"` in the tree is actually coded by the subfamily, `"Glossophaginae"`, in our association table.

```{r}
assoc[assoc$Bats=="Glossophaginae",1]<-"Glossophaga soricina"
```

```{r}
obj<-cophylo(bat.tree,virus.tree,assoc=assoc)
obj
```

We're done!

```{r}
plot(obj,link.lty="solid",link.type="curved",fsize=c(0.9,0.8))
```


